GLM Predictor Matrices

### 2020-12-14 workflow/scripts/predictor_matrices.Rmd
covariate matrix and analysis files europe6

```{r, echo = FALSE, results ='hide', message = FALSE}
library(knitr)
library(wpp2019)
library(tidyverse)
library(readr)
library(ggsci)
library(scales)
source("./scripts/utils.R")
source("./scripts/flight_data.R")
```

@@

- Eurostats, European data and statistics, transport data: <https://ec.europa.eu/eurostat/web/transport/data/database>
- United Nations Population Division, World Population Prospects 2019: R library `wpp2019, data(pop)`
- Distance

@@

fligh_data <- get_flightData(intraUE_data, extraUE_data, countries = demes)

fligh_data_3i <- fligh_data %>%
  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
  mutate(EPOCH = case_when(
    MONTH == 2 ~ 2,
    MONTH == 3 ~ 3,
    TRUE ~ 1)) %>%
  group_by(GEO, PARTNER, EPOCH) %>%
  summarise(N = n(),
         EVALUE = sum(DVALUE/N)) %>%
  ungroup %>%
  # Transformations
  mutate(LVALUE = log(EVALUE + 1),
         SLVALUE = scale(LVALUE)) %>%
  arrange(EPOCH, factor(GEO, levels = demes), factor(PARTNER, levels = demes))

#write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
```

```{r, echo = FALSE}
flight_matrices <- lapply(1:3, function (i) {
  fligh_data_3i %>%
  filter(EPOCH == i) %>%
  select(GEO, PARTNER, SLVALUE) %>%
  pivot_wider(names_from = "PARTNER", values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(.,2)) %>%
  column_to_rownames("GEO") %>%
  select(unname(demes))
})

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 3}
set_plotopts()
dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- dcolors[1:length(demes)]
names(dcolors) <- demes[sort(labels(demes))]
#dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
#names(dark_dcolors) <- names(dcolors)

# Fligh data plot
ggplot(fligh_data_3i, aes(x = EPOCH, y = EVALUE, color = PARTNER)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
  facet_grid(~GEO) +
  xlab("Epoch") +
  ylab("# Daily Flight Passengers") + 
  #theme(axis.title.x = element_blank()) + 
  scale_color_manual(values = dcolors, name = "Destination")
```

@@

```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name)
```

@@

A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across demes.

\tiny
```{r, echo=FALSE}
# TODO use long format, easy to combine later to export to xml, like with the flight data
# Population source matrix
pop_source <- matrix(1, 6, 6) * population$`2020`
diag(pop_source) <- NA
vslpop_source <- scale(log(c(pop_source)))
slpop_source <- matrix(vslpop_source, 6, 6)
colnames(slpop_source) <- demes[population$name]
rownames(slpop_source) <- demes[population$name]

# Population destination matrix
pop_dest <- sweep(matrix(1, 6, 6), 2, population$`2020`, `*`)
diag(pop_dest) <- NA
vslpop_dest <- scale(log(c(pop_dest)))
slpop_dest <- matrix(vslpop_dest, 6, 6)
colnames(slpop_dest) <- demes[population$name]
rownames(slpop_dest) <- demes[population$name]

knitr::kable(list(round(slpop_source, 2), round(slpop_dest, 2)), booktabs = TRUE, caption = "Population counts source (left) and destination (right)")
```

@@

## Distance between countries
Options:
Great circle distance between centroids of each country - Which distance for other european deme? Could be the average
Average pairwaise distance between airports - considers that only flight transportation matters, more work
Distance taking into account population distribution inside the country.
Minimum distance.

>>>>

```{r, echo = FALSE, results ='hide', message = FALSE}
library(knitr)
library(wpp2019)
library(tidyverse)
library(readr)
library(ggsci)
library(scales)
library(countrycode)
library(geosphere)
library(maps)
library(sf)
source("./scripts/utils.R")
source("./scripts/flight_data.R")
```

@@

- Eurostats, European data and statistics, transport data: <https://ec.europa.eu/eurostat/web/transport/data/database>
- United Nations Population Division, World Population Prospects 2019: R library `wpp2019, data(pop)`
- World map from the Natural Earth project (2013), loaded with R library `maps`.

@@

flight_df <- get_flightData(intraUE_data, extraUE_data, countries = demes)

flight_df_3e <- flight_df %>%
  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
  mutate(EPOCH = case_when(
    MONTH == 2 ~ 2,
    MONTH == 3 ~ 3,
    TRUE ~ 1),
    PRED = 1) %>%
  rename(SRC = GEO, DEST = PARTNER) %>%
  group_by(SRC, DEST, EPOCH, PRED) %>%
  summarise(N = n(),
         EVALUE = sum(DVALUE/N)) %>%
  ungroup %>%
  # Transformations
  mutate(LVALUE = log(EVALUE + 1),
         SLVALUE = scale(LVALUE)) %>%
  arrange(EPOCH, factor(SRC, levels = demes), factor(DEST, levels = demes))

#write(slv_flightMatrix_3e, "files/flight_matrix_3e.csv", sep = ",")
```

```{r, echo = FALSE}
flight_matrices <- lapply(1:3, function (i) {
  flight_df_3e %>%
  filter(EPOCH == i) %>%
  select(SRC, DEST, SLVALUE) %>%
  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(.,2)) %>%
  column_to_rownames("SRC") %>%
  select(all_of(unname(demes)))
})

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 3}
set_plotopts()
dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- dcolors[1:length(demes)]
names(dcolors) <- demes[sort(labels(demes))]

# Fligh data plot
ggplot(flight_df_3e, aes(x = EPOCH, y = EVALUE, color = DEST)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
  facet_grid(~SRC) +
  xlab("Epoch") +
  ylab("# Daily Flight Passengers") + 
  #theme(axis.title.x = element_blank()) + 
  scale_color_manual(values = dcolors, name = "Destination")
```

@@

```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name) %>%
  rename(POP = 2) %>%
  mutate(SLVALUE = scale(log(POP)))
pop_df <- population %>%
  mutate(n = 6, SRC = demes) %>%
  uncount(n) %>%
  mutate(DEST = rep(demes, 6),
         POP = ifelse(SRC == DEST, NA, POP),
         SLVALUE = ifelse(SRC == DEST, NA, SLVALUE)) %>%
  select(-name) %>%
  filter(!is.na(POP))
pop_df_3e <- bind_rows(replicate(3, pop_df %>% mutate(PRED = 2), simplify = FALSE)) %>% 
  mutate(EPOCH = rep(1:3, each = 30, length = 90)) %>%
  bind_rows(bind_rows(replicate(3, pop_df %>% 
                        arrange(factor(DEST, levels = demes),
                                factor(SRC, levels = demes)) %>%
                          mutate(TEMP = SRC,
                                 SRC = DEST,
                                 DEST = TEMP,
                                 PRED = 3),
                      simplify = FALSE)) %>%
  mutate(EPOCH = rep(1:3, each = 30, length = 90)))
```

@@

A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across epochs.

\tiny
```{r, echo=FALSE}
pop_matrices <- lapply(c("DEST", "SRC"), function (x) {
  pop_df %>%
  select(-POP) %>%
  pivot_wider(names_from = all_of(x), values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  mutate_at(1, ~factor(., levels = demes)) %>%
  arrange_at(1) %>%
  select(1, all_of(unname(demes)))
})

knitr::kable(pop_matrices, booktabs = TRUE, caption = "Population counts source (left) and destination (right)")
```

@@

## Geographical distance

```{r, echo = FALSE, warning = FALSE, message = FALSE}
map <- st_as_sf(map("world", plot = FALSE, fill = TRUE)) %>%
  mutate(country = ifelse(ID == "UK", "United Kingdom", as.character(ID)),
         continent = countrycode(sourcevar = country,
                             origin = "country.name",
                             destination = "continent")) %>%
  filter(country %in% labels(demes) | continent == "Europe") %>%
  mutate(deme = case_when(
    country %in% labels(demes) ~ country,
    TRUE ~ "OtherEuropean"))
map <- cbind(map, st_coordinates(st_centroid(map))) %>%
  rename(longitude = X, latitude = Y) %>%
  mutate(deme = demes[deme])

centroids <- data.frame(map) %>%
  select(country, deme, longitude, latitude)

dist_df <- centroids %>%
  full_join(centroids, by = character())
dist_df <- dist_df%>%
  mutate(DIST = distHaversine(select(dist_df, longitude.x, latitude.x),
                              select(dist_df, longitude.y, latitude.y))) %>%
  group_by(deme.x, deme.y) %>%
  summarise(DIST = mean(DIST), .groups = "drop") %>%
  filter(deme.x != deme.y) %>%
  rename(SRC = deme.x, DEST = deme.y) %>% 
  mutate(SLVALUE = scale(log(DIST))) %>%
  arrange(factor(SRC, levels = demes), factor(DEST, levels = demes))
dist_df_3e <- bind_rows(replicate(3, dist_df, simplify = FALSE)) %>%
  mutate(EPOCH = rep(1:3, each = 30, length = 90), PRED = 4)
```

Great circle pairwise distance between centroids of each country. Centroids are computed  from the Natural Earth project (the 1:50m resolution version) world map 2013. Constant across epochs. Symmetric.

\tiny
```{r, echo = FALSE}
dist_matrix <- dist_df %>%
  select(-DIST) %>%
  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  column_to_rownames("SRC") %>%
  select(all_of(unname(demes)))

knitr::kable(dist_matrix, booktabs = TRUE, caption = "Great circle pairwise distance between countries geographic centroid.")
```

\normalsize
The distance from one deme to the Other European deme is the average distance between the deme centroid and all the Other European countries centroids.

## Geographical distance

```{r, echo = FALSE, message =  FALSE, fig.align="center"}

ggplot() +
    geom_sf(data = map, aes(fill = deme), alpha = 0.7) + 
    geom_point(data = map %>% filter(country %in% labels(demes)), aes(x = longitude, y = latitude), shape = 4, size = 2) +
    coord_sf(xlim = c(-15, 120), ylim = c(25, 60), expand = FALSE) +
  scale_fill_manual(values = dcolors, na.value = "grey") +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

Map of Europe and part of Asia showing the demes included in the analysis (in color). The X represents the centroid of each deme. The centroids of each European country fromm the Other European deme are not drawed.

<!-- Other distance ptions: -->
<!-- Average pairwaise distance between airports - considers that only flight transportation matters, more work -->
<!-- Distance taking into account population distribution inside the country. CEPPI dataset -->
<!-- Minimum distance. -->

## Covariate list BEAST BDMM-Prime

The three predictors are saved in the following order: **Flight data, population count source, population count destination and geographical distance.**

Each of the predictor is composed of three matrices, one per epoch, with the first matrix being the most recent epoch, e.g. the first 30 values correponds to the fligh data for March.

Diagonal values are skipped.

Therefore the final size of the covariate list is of dimension $(30 \cdot 3) \cdot 4 = 360$.

```{r, echo = FALSE}
pred_3e <- bind_rows(flight_df_3e, pop_df_3e) %>%
  bind_rows(dist_df_3e) %>%
  arrange(PRED, desc(EPOCH), factor(SRC, levels = demes), factor(DEST, levels = demes)) %>%
  select(PRED, EPOCH, SRC, DEST, SLVALUE)

if (nrow(pred_3e) != 360) warning("Something went wrong, the size of the combined predictors is ", nrow(pred_3e))

mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(SLVALUE), ncol = 5, byrow = TRUE))
write_csv(mpred_3e, "./files/covariates_4p_3e.csv", col_names = FALSE)
```

\

Covariate list saved as `files/covariates_4p_3e.csv`
